Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-06-19 POPL2026/SlottedNotes.tex
slides wip

@@

>>>>

\usepackage{minted}

@@

\subsection{E-matching}

E-matching is a function $match(p,m * a, \beta, m_p)$ that takes in a pattern $p$, an applied-id $m*a$ to match against, a subtitution from variables to applied-ids $\beta$ and a renaming $m_p$ that maps slots of an e-graph to slots of the pattern.

\begin{listing}
\begin{minted}[linenos, fontsize=\small, breaklines]{rust}
  struct State {
    // uses egraph slots.
    partial_subst: Subst, // type Subst = HashMap<String, AppliedId>

    // maps from the pattern slots to the e-graph slots.
    partial_slotmap: SlotMap, //type SlotMap = HashMap<Slot,Slot>
  }

  pub fn ematch_all<L: Language, N: Analysis<L>>(
    eg: &EGraph<L, N>,
    pattern: &Pattern<L>,
) -> Vec<Subst> {
    let mut out = Vec::new();
    for i in eg.ids() {
        let i = eg.mk_sem_identity_applied_id(i); // applies identity renaming to an id to turn it into AppliedId
        out.extend(
            ematch_impl(pattern, State::default(), i, eg) // this is `match'
                .into_iter()
                .map(final_subst),
        );
    }
    out
}

\end{minted}
\end{listing}

\begin{listing}
\begin{minted}[linenos, fontsize=\small, breaklines]{rust}
fn ematch_impl<L: Language, N: Analysis<L>>(
    pattern: &Pattern<L>,
    st: State,
    i: AppliedId,
    eg: &EGraph<L, N>,
) -> Vec<State> {
    match &pattern {
        Pattern::PVar(v) => {
            let mut st = st;
            if let Some(j) = st.partial_subst.get(v) {
                if !eg.eq(&i, j) {
                    return Vec::new();
                }
            } else {
                st.partial_subst.insert(v.clone(), i);
            }
            vec![st]
        }
        Pattern::ENode(n, children) => {
            let mut out = Vec::new();
            for nn in eg.enodes_applied(&i) { //traverses e_nodes of the e-class
                let d = std::mem::discriminant(n);
                let dd = std::mem::discriminant(&nn);
                if d != dd {
                    continue;
                };

                //takes the e_node that matches 'n'.
                ematch_node(&st, eg, &n, children, &mut out, &nn);
            }
            out
        }
        Pattern::Subst(..) => panic!(),
    }
}

\end{minted}
\end{listing}

\begin{listing}
\begin{minted}[linenos, fontsize=\small, breaklines]{rust}
fn ematch_node<L: Language, N: Analysis<L>>(
    st: &State,
    eg: &EGraph<L, N>,
    n: &L, // from the pattern
    children: &[Pattern<L>],
    out: &mut Vec<State>,
    nn: &L, //candidate e_node from e-class
) {
    'nodeloop: for n2 in eg.get_group_compatible_weak_variants(&nn) {
        if CHECKS {
            assert_eq!(&nullify_app_ids(n), n);
        }

        let clear_n2 = nullify_app_ids(&n2);
        // We can use weak_shape here, as the inputs are nullified
        // i.e. they only have id0() without slot args, so there are no permutations possible.
        let (n_sh, _) = n.weak_shape();
        let (clear_n2_sh, _) = clear_n2.weak_shape();
        if n_sh != clear_n2_sh {
            continue 'nodeloop;
        } //checks if the shapes are equal, for example lambda x . f and lambda y . f have the same shape

        let mut st = st.clone();

        for (x, y) in clear_n2
            .all_slot_occurrences()
            .into_iter()
            .zip(n.all_slot_occurrences().into_iter())
        {
          // checks if mapping m_p is not ambiguous
            if !try_insert_compatible_slotmap_bij(x, y, &mut st.partial_slotmap) {
                continue 'nodeloop;
            }
        }

        let mut acc = vec![st];
        // recursively matches children
        for (sub_id, sub_pat) in n2.applied_id_occurrences().into_iter().zip(children.iter()) {
            let mut next = Vec::new();
            for a in acc {
                next.extend(ematch_impl(sub_pat, a, sub_id.clone(), eg));
            }
            acc = next;
        }

        out.extend(acc);
    }
}

\end{minted}
\end{listing}

### 2025-06-19 slides/EGRAPHS_2024/slides.tex
slides wip

\input{../sample.tikzstyles}
\input{../hypergraph.tikzstyles}
\input{../hypergraph.tikzdefs}

@@

% \setsansfont{Ubuntu}
% \setmonofont{Ubuntu Mono}

@@

\title{Equivalence Hypergraphs: E-Graphs for Monoidal Theories} 
\date{June 24, 2024} % Use metropolis theme 
\author[Aleksei Tiurin]{$\text{Chris Barrett}^{\dagger}$, $\textbf{Aleksei Tiurin}^{\ddagger}$, $\text{Dan R. Ghica}^{\ddagger}$}
\institute{$^{\dagger}$University of Oxford, $^{\ddagger}$University of Birmingham} 
\begin{document}

@@

\begin{frame}{}
    \[
        \scalebox{0.6}{
        \tikzfig{../figures/categorical-semantics/egraph-strings-equations}
        }
    \]
\end{frame}

@@

    That is, if we consider the generators (the inputs flow from bottom to top to follow the e-graphs presentation)
    \[
	\scalebox{0.7}{
  	 \tikzfig{../figures/categorical-semantics/Cartesian-equipment}
	}
    \]

@@

    \begin{minipage}{0.4\linewidth}
        \[
            \scalebox{0.55}{
            \tikzfig{../figures/categorical-semantics/Cartesian-theory}	
            }
            \]
    \end{minipage}
    \hfill
    \begin{minipage}{0.5\linewidth}
        \[
            \scalebox{0.55}{
            \tikzfig{../figures/categorical-semantics/Cartesian-naturality}
            }
        \]
    \end{minipage}

@@

    \[
    \scalebox{0.5}{
    \tikzfig{figures/e-graphs-translation-3}
    }
    \]\footnote{[1]Willsey, M., Nandi, C., Remy Wang, Y., Flatt, O., Tatlock, Z., and Panchekha, P., “egg: Fast and Extensible Equality Saturation”, 2020.}
    Can be rendered as
    \[
        % \hspace{1.3cm}
        \scalebox{0.38}{
        \tikzfig{../figures/categorical-semantics/egraph-translation-step-by-step-a-b}
        }
    \]
\end{frame}

@@

\begin{frame}
    \[
        % \hspace{1.3cm}
        \scalebox{0.4}{
        \tikzfig{../figures/categorical-semantics/e-graph-substitution}
        }
    \]
    \tiny{[2]\href{https://pldi23.sigplan.org/details/egraphs-2023-papers/12/Optimizing-Beta-Reduction-in-E-Graphs}{Optimizing-Beta-Reduction-in-E-Graphs}}
    \[
        % \hspace{1.3cm}
        \scalebox{0.35}{
        \tikzfig{../figures/categorical-semantics/e-string-substitution-2}
        }
    \]
\end{frame}

>>>>

\input{../../sample.tikzstyles}
\input{../../hypergraph.tikzstyles}
\input{../../hypergraph.tikzdefs}

@@

\setsansfont{Ubuntu}
\setmonofont{Ubuntu Mono}

@@

\title{Equivalence Hypergraphs: E-Graphs for Monoidal Theories} 
\date{June 24, 2024} % Use metropolis theme 
\author[Aleksei Tiurin]{$\textbf{Aleksei Tiurin}^{\dagger}$,$\text{Chris Barrett}^{\dagger}$,$\text{Dan R. Ghica}^{\dager,\ddagger}$, $\text{Nick Hu}^{\dagger}$}
\institute{$^{\dagger}$University of Birmingham, $^{\dagger}$ Huawei Central Software Institute}
\begin{document}

@@

\begin{frame}{}
    \[
        \scalebox{0.6}{
        \tikzfig{../../figures/categorical-semantics/egraph-strings-equations}
        }
    \]
\end{frame}

@@

    That is, if we consider the generators (the inputs flow from bottom to top to follow the e-graphs presentation)
    \[
	\scalebox{0.7}{
  	 \tikzfig{../../figures/categorical-semantics/Cartesian-equipment}
	}
    \]

@@

    \begin{minipage}{0.4\linewidth}
        \[
            \scalebox{0.55}{
            \tikzfig{../../figures/categorical-semantics/Cartesian-theory}	
            }
            \]
    \end{minipage}
    \hfill
    \begin{minipage}{0.5\linewidth}
        \[
            \scalebox{0.55}{
            \tikzfig{../../figures/categorical-semantics/Cartesian-naturality}
            }
        \]
    \end{minipage}

@@

    \[
    \scalebox{0.5}{
    \tikzfig{figures/e-graphs-translation-3}
    }
    \]\footnote{[1]Willsey, M., Nandi, C., Remy Wang, Y., Flatt, O., Tatlock, Z., and Panchekha, P., “egg: Fast and Extensible Equality Saturation”, 2020.}
    Can be rendered as
    \[
        % \hspace{1.3cm}
        \scalebox{0.38}{
        \tikzfig{../../figures/categorical-semantics/egraph-translation-step-by-step-a-b}
        }
    \]
\end{frame}

@@

\begin{frame}
    \[
        % \hspace{1.3cm}
        \scalebox{0.4}{
        \tikzfig{../../figures/categorical-semantics/e-graph-substitution}
        }
    \]
    \tiny{[2]\href{https://pldi23.sigplan.org/details/egraphs-2023-papers/12/Optimizing-Beta-Reduction-in-E-Graphs}{Optimizing-Beta-Reduction-in-E-Graphs}}
    \[
        % \hspace{1.3cm}
        \scalebox{0.35}{
        \tikzfig{../../figures/categorical-semantics/e-string-substitution-2}
        }
    \]
\end{frame}
